Longitudinal Manifold Estimation

### 2023-02-14 paper/main.tex
introduction draft

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al (2021), Poewe et al (2017)). Frequently, interest centers around changes in one or more substructures. For example, it is common to observe atrophy in the hippocampus of those with Alzheimer's disease or cognitive impairment from other causes. Image segmentation allows for the extraction and analysis of these subcortical structures.

@@

Ultimately, high levels of variability are present at the subcortical level regardless of the segmentation approach used, but may be particularly influential when using automated segmentation methods. To address these elevated noise levels, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time. This approach, which extends the principal manifold estimation algorithm introduced by Meng and Eloyan (2021) for a longitudinal setting, includes a regularization parameter to limit the effects of between time-point variability.

\section{Introduction - Old}

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold. These methods have become increasingly popular for medical imaging applications, where assuming an underlying manifold structure is helpful for modeling certain anatomical shapes. In these settings, interacting with a structure in a low-dimensional form may be more intuitive than in a higher-dimensional space. For example, Yue et al. (2016) seeks to compute a parametric representation of the corpus callosum. A number of methods have been developed for manifold learning, including Isomap (Tenenbaum et al. 2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003). Meng and Eloyan (2021) proposed a principal manifold framework that resolved several limitations of previously developed extensions of the principal curves approach introduced by Hastie and Stuetzle (1989).

In many medical imaging situations, understanding how an anatomical structure changes over time could provide meaningful information into the function of that structure, or medical conditions that affect that function. In situations where the structure is better represented by a low-dimensional manifold, interest carries over to understanding how the underlying manifold changes longitudinally as well.

While the methods given above provide a variety of approaches to retrieving a low-dimensional manifold from high-dimensional data at a single time point, they are ill-equipped to reach meaningful estimates across multiple time points. This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension-reduction process. This treatment of the time dimension causes time to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time.

Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in Meng and Eloyan (2021) to estimate an appropriate principal manifold at each given time point. We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold.

>>>>

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease and Parkinson's disease (Knopman et al (2021), Poewe et al (2017)). Frequently, interest centers around longitudinal changes in one or more substructures. For example, it is common to observe atrophy in the hippocampus of those with Alzheimer's disease or cognitive impairment from other causes. Image segmentation allows for the extraction and analysis of these subcortical structures.

@@

Ultimately, high levels of variability are present at the subcortical level regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods. In this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time. Manifold learning describes a set of statistical approaches to modeling high-dimensional data that are assumed to lie along a low-dimensional manifold. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive than in a higher-dimensional space. For example, Yue et al. (2016) seeks to compute a parametric representation of the corpus callosum. To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by Meng and Eloyan (2021).

Previously introduced methods for manifold learning, such as Isomap (Tenenbaum, et al. 2000), locally linear embedding (Roweis and Saul, 2000), and Laplacian eigenmaps (Belkin and Niyogi, 2003), are ill-equipped to reach meaningful estimates across multiple time points. This is because these methods consider time as an additional dimension in the data, meaning they include time in the dimension-reduction process. This treatment of the time dimension causes time to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time.

Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in Meng and Eloyan (2021) to estimate an appropriate principal manifold at each given time point. We then smooth over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. This smoothing process conducted between longitudinal observations mitigates the effects of
